Basics of Hardware Cache Optimization

### 2017-05-24 tex/no-prefetch.tex
Add awesome plot, expand content, tweak stuff

\section{Reading Information About the CPU}
\label{app:cpuinfo}
% https://stackoverflow.com/q/7281699
% https://unix.stackexchange.com/q/167038
% https://superuser.com/q/55776
There are many ways to display information about the processor(s) the operating system is
running on.  Among others, the \mintinline{text}{lscpu(1)} and
\mintinline{text}{getconf(1)} programs and the \mintinline{text}{/proc/cpuinfo}
pseudo-file on Linux.  This is \alts{how I checked, what I used to check} my CPU's cache
line sizes, for example:
\begin{minted}[gobble=3]{bash}
   $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
   64
   64
\end{minted}
%stopzone

% Performance Analysis
% Run Time Analysis
% Timing Methodology
% Profiling Methodology
% Profiling Pitfalls
% Run Time Measuring
% Benchmarking Methodology
% Monitoring
\section{Profiling Methodology}
% https://en.wikipedia.org/wiki/Profiling_(computer_programming)
% https://en.wikipedia.org/wiki/Processor_affinity
% On `/usr/bin/time` for benchmarks: <http://stackoverflow.com/q/9006596>.
% Andrei Alexandrescu - Writing Fast Code: <https://youtu.be/vrfYLlR8X8k>.
Plenty of things can go wrong when measuring \alts{execution, run} times.
\begin{itemize}
   \item CPU frequency scaling and boosting
      % Not a problem when measuring cycles.
   \item Interrupts
      % Taking the minimum helps.
   \item CPU jumping
      % See <https://youtu.be/vrfYLlR8X8k?t=33m34s>.
   \item Context switches
   \item Warmth of cache
      % See <https://youtu.be/vrfYLlR8X8k?t=41m00s>.
   % \item ...
\end{itemize}
Some effects will \alts{add noise to, pollute} the results or render them irreproducible,
others will invalidate them completely.

\begin{comment}
\section{Disabling the Prefetcher}
My initial results from from running \cref{lst:array-sum} were quite different.  See TODO.

TODO: I installed \texttt{msr-tools} and OProfile.
% This seems to work.  It did affect performance.
\begin{minted}[gobble=3]{bash}
   $ sudo modprobe msr
   $ sudo rdmsr -a 0xC0011022
   600001800000000
   600001800000000
   $ sudo wrmsr -a 0xC0011022 0x600001800002000
   $ sudo rdmsr -a -f '13:13' 0xC0011022
   1
   1
   $ sudo rdmsr -a 0xC0011022
   600001800002000
   600001800002000
\end{minted}
%stopzone
I think resetting the MSR to enable hardware prefetches doesn't work.  To get the same
benchmmark results as before tweaking it, I have to reboot.
\end{comment}

>>>>

\section{Disabling the Prefetcher}
My initial results from from running \cref{lst:array-sum} were quite different.  See TODO.

TODO: I installed \texttt{msr-tools} and OProfile.
% This seems to work.  It did affect performance.  See
% <https://community.amd.com/thread/180725>.
\begin{minted}[gobble=3]{bash}
   $ sudo modprobe msr
   $ sudo rdmsr -a 0xC0011022
   600001800000000
   600001800000000
   $ sudo wrmsr -a 0xC0011022 0x600001800002000
   $ sudo rdmsr -a -f '13:13' 0xC0011022
   1
   1
   $ sudo rdmsr -a 0xC0011022
   600001800002000
   600001800002000
   $ sudo ocount -e PREFETCH_INSTRUCTIONS_DISPATCHED ./a.out
\end{minted}
%stopzone
I think resetting the MSR to enable hardware prefetches doesn't work.  To get the same
benchmmark results as before tweaking it, I have to reboot.

### 2017-05-24 tex/types.tex
Add awesome plot, expand content, tweak stuff

>>>>

My laptop's AMD E-450 CPU has cores with \alts{an \gls{l1d} cache of 32 KiB, 32 KiB of
\gls{l1d} cache} and a unified \gls{l2} cache of 512 KiB each.%
\footnote{\Cref{app:cpuinfo} explains how to obtain this information.}
We can both verify these sizes and get \alts{a reasonably good measure, a rough measure,
an approximation} of the access times by profiling
\cref{lst:access-times}
% the \alts{program, listing} below
for different values of \mintinline{text}{SIZE}.%
% FIXME: the label is wrong: should be "Appendix" but is "Section".
\footnote{\Cref{app:cycles} details how.}
% It reads random memory addresses
% The program randomly reads
% locations
% The program repeatedly \alts{reads, accesses} random elements
This program repeatedly \alts{reads, accesses} elements
from \alts[2]{an array of the configured size, a thusly sized array}
in random \alts{order, sequence, succession}.
To do this
\alts{with minimal overhead, efficiently}, the array \alts{is first set up, acts} as a
circular, singly linked list where every element except the last \alts{points to a, has a}
random successor.  When compiled with \mintinline{text}{-DBASELINE}, only this
initialization is done.
\Cref{fig:access-times} shows the deltas of CPU cycles used when and when not having
defined \mintinline{text}{BASELINE}.
I divided by \mintinline{text}{N} to get the cycles spend on each loop iteration.

% XXX: hacks!  Use the figure environment so that LaTeX won't display this listing after
% the plot showing the results of profiling it.  "LaTeX [only] keeps all floats of the
% same *type* in order" [1].  Use `\captionof` to label the listing correctly as a
% listing.
%
% [1]: https://tex.stackexchange.com/q/127742/#comment290982_127744
\begin{comment}
   \begin{figure}
      \inputminted[firstline=9]{c}{access-times/access-times.c}
      \captionof{listing}{TODO}
      \label{lst:access-times}
   \end{figure}
\end{comment}

% \newenvironment{code}{\captionsetup{type=listing}}{}
% \begin{code}
%    \inputminted[firstline=9]{c}{access-times/access-times.c}
%    \caption{TODO}
%    \label{lst:access-times}
% \end{code}

% Actually, don't use a float.  The listing should be allowed to span multiple pages which
% floats aren't.
%
% [1]: https://tex.stackexchange.com/q/14522/#comment484569_75880
% [2]: https://tex.stackexchange.com/q/175650
%      "How to allow page break inside a float environment?"
% [3]: https://tex.stackexchange.com/q/12428
\inputminted[firstline=9]{c}{access-times/access-times.c}
\captionof{listing}{}
% \captionof{listing}{Random Reads\label{lst:access-times}}
\label{lst:access-times}

% https://tex.stackexchange.com/a/82473
% \global\csname @topnum\endcsname 0

% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
% interrupting the source code listing.
\begin{figure}[hbp]
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               ticks={major={at={2048 as 2, 8192 as 8, 32768 as \textbf{32},
                                 131072 as 128, 524288 as \textbf{512}, 2097152 as 2048,
                                 8388608 as 8192, 33554432 as 32768, 134217728 as
                                 % $128\cdot 2^{10}$}}},
                                 % $2^{17}$}}},
                                 131072}}},
               grid={at={32768, 524288}},
               label={Array Size (KiB)}, length=0.8\textwidth},
       y axis={include value=0, label={Cycles / Iteration}, length=6cm, grid=at ticks},
       % y axis={logarithmic,
       %         ticks={major={at={3, 15, 23, 190}}},
       %         grid={at={15, 23}},
       %         label={Cycles / Element}, length=6cm},
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         data [read from file=access-times/access-times.csv, separator=\space];
   \end{tikzpicture}
   % \caption{Sequential Read Access}
   % \caption{Access Times for Random Reads (\Cref{lst:access-times})}
   \caption{Access Times for Random Reads}
   \label{fig:access-times}
\end{figure}
